Assessment and Optimization of Nowcast Measures for Trend Detection

### 2023-12-19 paper.tex
Add new weighted estimator; remove whitespace in folder

\newcommand{\lag}[1][l]{\Delta_{#1}}
\DeclarePairedDelimiter{\abs}\lvert\rvert
\DeclareMathOperator{\sign}{sign}
\newcommand{\tmax}{\bar{t}}
\newcommand{\ind}[1]{\mathbbm{1}\{#1\}}
\newcommand{\ydiff}{D y}
\newcommand{\ydifft}{Dy^\star}
\newcommand{\xdiff}{Dx}
\newcommand{\xdifft}{Dx^\star}
\newcommand{\Ydiff}{DY}
\newcommand{\Xdiff}{DX}
\newcommand{\Prob}[1]{P(#1)}
\newcommand{\mprob}{\tilde{m}}
\newcommand{\Ber}{\text{Ber}}
\newcommand{\SBer}{\text{SBer}}

@@

@@

% \begin{itemize}
%   \item Quadratic/Brier Score for $k$ possible outcomes \hl{Erst mal rauslassen; wäre anderes setting: probabilistischer Nowcast}: \begin{equation}
%   S((p_1, \dots, p_k), i) = 2 p_i - \sum_{l=1}^k p_i^2 - 1;
% \end{equation}
% for dichotomous outcomes ($p_1 = \hat{P}(\ydiff_t > 0)$: probability for $\ydiff_t > 0$):
% \begin{equation}
%   S(p_1, \ydiff_t) = p_1 \ind{\ydiff_t > 0} + (1 - p_1) (1 - \ind{\ydiff_t > 0}) + p_1^2 + (1-p_1)^2 - 1
% \end{equation}
% \end{itemize}
\item Probabilistic versions of non-probabilistic measures; small sample characteristic of $m_{acc}$ can be driven by random effects near the axis; therefore probabilistic \enquote{correction}:
\begin{itemize}
  \item Assume an additive error decomposition for both observation and nowcast to account for non-systematic and short-term (i.e., intra-day) effects or for effects on inaccurate representation of $y_{t-1}$ in the computation of $Dx_t$ (\hl{this, no additional complexity}):
  	\begin{align}\label{additive error decomposition}
  		\ydiff_t &= \ydifft + \varepsilon_t^y \\
  		\xdiff_t &= \xdifft + \varepsilon_t^x
	\end{align}
 \item \hl{Wouldn't it be more natural to model the original time series with a noise term?} (Qualitatively, it would yield the same results)
 \begin{align}
     y_t &= y_t^\star + \epsilon_t^y \\
     x_t &= x_t^\star + \epsilon_t^x \\
     \ydiff_t &= y_t^\star + \epsilon_t^y - (y_{t-1}^\star + \epsilon_{t-1}^y) = y_t^\star - y_{t-1}^\star + \underbrace{\epsilon_t^y - \epsilon_{t-1}^y}_{\sim N(0, \sqrt{2} \sigma_y)} \\
     \xdiff_t &= x_t^\star + \epsilon_t^x - (x_{t-1}^\star + \epsilon_{t-1}^x) = x_t^\star - x_{t-1}^\star + \underbrace{\epsilon_t^x - \epsilon_{t-1}^x}_{\sim N(0, \sqrt{2} \sigma_x)}
 \end{align}
 Advantage of the approach: For different lags, no new standard deviation has to be estimated.\\
 Disadvantage: Theoretically, there is serial correlation of the errors. (Wie behandle ich die Autokorrelation; wie rausrechnen)
\item  Replace indicators in the above formulations by the corresponding probabilities:
\begin{align}
		\mprob_{\text{acc}} &= \frac{\sum_{t=T-w}^T \Prob{ \ydifft_t > 0, \xdifft_t > 0} + \sum_{t=T-w}^T \Prob{\ydifft_t < 0, \xdifft_t < 0}}{T-w}  \\
   \mprob_{\text{py}} &= \frac{\sum_{t=T-w}^T \Prob{\ydifft_t > 0, \xdifft_t > 0}}{\sum_{t=T-w}^T \Prob{\ydifft_t > 0}} \\
    \mprob_{\text{px}} &= \frac{\sum_{t=T-w}^T \Prob{\ydifft_t > 0, \xdifft_t > 0}}{\sum_{t=T-w}^T \Prob{\xdifft_t > 0}} \\
    \mprob_{\text{ny}} &= \frac{\sum_{t=T-w}^T \Prob{\ydifft_t < 0, \xdifft_t < 0}}{\sum_{t=T-w}^T \Prob{\ydifft_t < 0}} \\
    \mprob_{\text{px}} &= \frac{\sum_{t=T-w}^T \Prob{\ydifft_t < 0, \xdifft_t < 0}}{\sum_{t=T-w}^T \Prob{\xdifft_t < 0}} 
\end{align} 
\item Simple error model:
  \begin{align}
	  \varepsilon_Y \sim N(0, \sigma_Y) \\
	  \varepsilon_X \sim N(0, \sigma_X)
  \end{align}
  yields, e.g.,
  	\begin{equation}
  		\mprob_{\text{acc}} = \frac{1}{T-w} \sum_{t=T-w}^T  \big( 1 - \Phi_{\ydiff_t, \sigma_y}(0) - \Phi_{\xdiff_t, \sigma_x} (0) + 2 \Phi_{\xdiff_t,\sigma_x}( 0)\Phi_{\ydiff_t,\sigma_y}( 0) \big), 
	\end{equation}
	where $\Phi$ denotes a (possibly multivariate) normal distribution
\end{itemize}
\item For $\sigma_x \sigma_y \rightarrow 0$, the probabilistic measures approach the count-based ones
\item For $\sigma_x \sigma_y \rightarrow \infty$, the probabilistic accuracy approaches   0.5 as all probabilities are approximately 0.5
\item Lag-$l$-measures can easily be obtained by using the $D^l$-difference, i.e., $D^l y_t = y_t - y_{t-l}$
\end{itemize}

@@

\begin{figure}
  \centering
    \begin{subfigure}{0.48\textwidth}
        \includegraphics[width=\textwidth]{plots/std estimation/std_estimation_boxplot_25.pdf}
    \end{subfigure}
    \begin{subfigure}{0.48\textwidth}
        \includegraphics[width=\textwidth]{plots/std estimation/std_estimation_boxplot_50.pdf}
    \end{subfigure}

    \begin{subfigure}{0.48\textwidth}
        \includegraphics[width=\textwidth]{plots/std estimation/std_estimation_boxplot_100.pdf}
    \end{subfigure}
    \begin{subfigure}{0.48\textwidth}
        \includegraphics[width=\textwidth]{plots/std estimation/std_estimation_boxplot_1000.pdf}
    \end{subfigure}

>>>>

\newcommand{\lag}[1][l]{\Delta_{#1}}
\DeclarePairedDelimiter{\abs}\lvert\rvert
\DeclareMathOperator{\sign}{sign}
\newcommand{\tmax}{\bar{t}}
\newcommand{\ind}[1]{\mathbbm{1}\{#1\}}
\newcommand{\ydiff}{D y}
\newcommand{\ydifft}{Dy^\star}
\newcommand{\xdiff}{Dx}
\newcommand{\xdifft}{Dx^\star}
\newcommand{\Ydiff}{DY}
\newcommand{\Xdiff}{DX}
\newcommand{\Prob}[1]{P(#1)}
\newcommand{\mprob}{\tilde{m}}
\newcommand{\Ber}{\text{Ber}}
\newcommand{\SBer}{\text{SBer}}
\newcommand{\cond}{\:\lvert\:}

@@

\hl{Idea: Weighted Indicator, accounting for the likeliness of a point's location being driven by noise}

\begin{itemize}
  \item Weights $\omega_t$ for $(\ydiff_t, \xdiff_t)$ could be determined by error model and corresponds to the maximal likeliness of the point belonging to the different quadrants ($Q_1, \dots, Q_4$); thus, point that have less explanatory power for assessing the trending ability becomes less weight:
 	\begin{equation}
  		\omega_t = \max_{i \in [4]} {P\big( (\ydifft_t, \xdifft_t) \in Q_i \cond \ydiff_t, \xdiff_t \big)} 
	\end{equation}
	\item E.g., the accuracy becomes
	\begin{equation}
  m_{acc, \omega} = \frac{\sum_{t=T-w}^T \omega_t \ind{\ydiff_t \cdot \xdiff_t > 0} }{\sum_{t=T-w}^T \omega_t}
\end{equation}
	\item Thus, $\omega_t \in [0.25, 1]$. Maybe it would make sense to \enquote{ignore} values without explanatory power and use $\omega_t^\star = 4/3 \ \omega_t \in [0, 1]$ \hl{to discuss}
	\item Figure~\ref{fig:comparison_estimators} shows the resulting estimates for the simulation study comparing the pure indicator, the probability-version and the weighted indicator estimation; the data is generated (in a very simpel model) by
	\begin{align}
		\ydifft_t &\sim N(0, 2) \\
		\xdifft_t &= (2 b_t - 1) \cdot \ydiff_t , b_t \sim \Ber (k) \\
		\ydiff_t &= \ydifft_t + \varepsilon_t^y, \varepsilon_t^y \sim N(0, 0.1) \\
		\xdiff_t &= \xdifft_t + \varepsilon_t^x, \varepsilon_t^x \sim N(0, 0.1) \\
	\end{align}
\end{itemize}

\begin{figure}
  \centering
	\includegraphics{plots/unbiasedness/boxplot_k_0.1.pdf}
	\includegraphics{plots/unbiasedness/boxplot_k_0.3.pdf}
	\includegraphics{plots/unbiasedness/boxplot_k_0.5.pdf}
	\includegraphics{plots/unbiasedness/boxplot_k_0.7.pdf}
	\includegraphics{plots/unbiasedness/boxplot_k_0.9.pdf}			
  \caption{Simulations results comparing the three types of estimators for the accuracy in a trending setting}
  \label{fig:comparison_estimators}
\end{figure}

@@

% \begin{itemize}
%   \item Quadratic/Brier Score for $k$ possible outcomes \hl{Erst mal rauslassen; wäre anderes setting: probabilistischer Nowcast}: \begin{equation}
%   S((p_1, \dots, p_k), i) = 2 p_i - \sum_{l=1}^k p_i^2 - 1;
% \end{equation}
% for dichotomous outcomes ($p_1 = \hat{P}(\ydiff_t > 0)$: probability for $\ydiff_t > 0$):
% \begin{equation}
%   S(p_1, \ydiff_t) = p_1 \ind{\ydiff_t > 0} + (1 - p_1) (1 - \ind{\ydiff_t > 0}) + p_1^2 + (1-p_1)^2 - 1
% \end{equation}
% \end{itemize}
\item Probabilistic versions of non-probabilistic measures; small sample characteristic of $m_{acc}$ can be driven by random effects near the axis; therefore probabilistic \enquote{correction}:
\begin{itemize}
  \item Assume an additive error decomposition for both observation and nowcast to account for non-systematic and short-term (i.e., intra-day) effects or for effects on inaccurate representation of $y_{t-1}$ in the computation of $Dx_t$ (\hl{this, no additional complexity}):
  	\begin{align}\label{additive error decomposition}
  		\ydiff_t &= \ydifft + \varepsilon_t^y \\
  		\xdiff_t &= \xdifft + \varepsilon_t^x
	\end{align}
 \item Wouldn't it be more natural to model the original time series with a noise term? (Qualitatively, it would yield the same results)
 \begin{align}
     y_t &= y_t^\star + \epsilon_t^y \\
     x_t &= x_t^\star + \epsilon_t^x \\
     \ydiff_t &= y_t^\star + \epsilon_t^y - (y_{t-1}^\star + \epsilon_{t-1}^y) = y_t^\star - y_{t-1}^\star + \underbrace{\epsilon_t^y - \epsilon_{t-1}^y}_{\sim N(0, \sqrt{2} \sigma_y)} \\
     \xdiff_t &= x_t^\star + \epsilon_t^x - (x_{t-1}^\star + \epsilon_{t-1}^x) = x_t^\star - x_{t-1}^\star + \underbrace{\epsilon_t^x - \epsilon_{t-1}^x}_{\sim N(0, \sqrt{2} \sigma_x)}
 \end{align}
 Advantage of the approach: For different lags, no new standard deviation has to be estimated.\\
 Disadvantage: Theoretically, there is serial correlation of the errors. (Wie behandle ich die Autokorrelation; wie rausrechnen)
\item  Replace indicators in the above formulations by the corresponding probabilities:
\begin{align}
		\mprob_{\text{acc}} &= \frac{\sum_{t=T-w}^T \Prob{ \ydifft_t > 0, \xdifft_t > 0} + \sum_{t=T-w}^T \Prob{\ydifft_t < 0, \xdifft_t < 0}}{T-w}  \\
   \mprob_{\text{py}} &= \frac{\sum_{t=T-w}^T \Prob{\ydifft_t > 0, \xdifft_t > 0}}{\sum_{t=T-w}^T \Prob{\ydifft_t > 0}} \\
    \mprob_{\text{px}} &= \frac{\sum_{t=T-w}^T \Prob{\ydifft_t > 0, \xdifft_t > 0}}{\sum_{t=T-w}^T \Prob{\xdifft_t > 0}} \\
    \mprob_{\text{ny}} &= \frac{\sum_{t=T-w}^T \Prob{\ydifft_t < 0, \xdifft_t < 0}}{\sum_{t=T-w}^T \Prob{\ydifft_t < 0}} \\
    \mprob_{\text{px}} &= \frac{\sum_{t=T-w}^T \Prob{\ydifft_t < 0, \xdifft_t < 0}}{\sum_{t=T-w}^T \Prob{\xdifft_t < 0}} 
\end{align} 
\item Simple error model:
  \begin{align}
	  \varepsilon_Y \sim N(0, \sigma_Y) \\
	  \varepsilon_X \sim N(0, \sigma_X)
  \end{align}
  yields, e.g.,
  	\begin{equation}
  		\mprob_{\text{acc}} = \frac{1}{T-w} \sum_{t=T-w}^T  \big( 1 - \Phi_{\ydiff_t, \sigma_y}(0) - \Phi_{\xdiff_t, \sigma_x} (0) + 2 \Phi_{\xdiff_t,\sigma_x}( 0)\Phi_{\ydiff_t,\sigma_y}( 0) \big), 
	\end{equation}
	where $\Phi$ denotes a (possibly multivariate) normal distribution
\end{itemize}
\item For $\sigma_x \sigma_y \rightarrow 0$, the probabilistic measures approach the count-based ones
\item For $\sigma_x \sigma_y \rightarrow \infty$, the probabilistic accuracy approaches   0.5 as all probabilities are approximately 0.5
\item Lag-$l$-measures can easily be obtained by using the $D^l$-difference, i.e., $D^l y_t = y_t - y_{t-l}$
\end{itemize}

@@

\begin{figure}
  \centering
    \begin{subfigure}{0.48\textwidth}
        \includegraphics[width=\textwidth]{plots/std_estimation/std_estimation_boxplot_25.pdf}
    \end{subfigure}
    \begin{subfigure}{0.48\textwidth}
        \includegraphics[width=\textwidth]{plots/std_estimation/std_estimation_boxplot_50.pdf}
    \end{subfigure}

    \begin{subfigure}{0.48\textwidth}
        \includegraphics[width=\textwidth]{plots/std_estimation/std_estimation_boxplot_100.pdf}
    \end{subfigure}
    \begin{subfigure}{0.48\textwidth}
        \includegraphics[width=\textwidth]{plots/std_estimation/std_estimation_boxplot_1000.pdf}
    \end{subfigure}
